Safe Transferable Regions

### 2017-01-13 paper/broom.tex
added alloc context to constraint generation rules
type inference minor updates
Formatting edits.
fixed typos in constraint generation rules
Edits to the intro.
Sec.3 edits complete.

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Gowthan Kaki}
\author[2]{G. Ramalingam}
\author[3]{Kapil Vaswani}
\author[4]{Dimitrios Vytiniotis}
\affil[1]{Purdue University}
\affil[2]{Microsoft Research}
\affil[3]{Microsoft Research}
\affil[4]{Microsoft Research}
\authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

% \subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Memory management, regions, type system, type inference}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{ECOOP 2017}
\EventAcronym{ECOOP}
\EventYear{2017}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\renewcommand{\rgn}{\pi}
\input{type-inference}
\renewcommand{\rgn}{r}

@@

@@

\input{appendix}

>>>>

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author{}
% \author[1]{Gowthan Kaki}
% \author[2]{G. Ramalingam}
% \author[2]{Kapil Vaswani}
% \author[2]{Dimitrios Vytiniotis}
% \affil[1]{Purdue University}
% \affil[2]{Microsoft Research}
% \authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

% \Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

% \subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Memory Safety, Formal Methods, Type System, Type Inference,
Regions}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{ECOOP 2017}
\EventAcronym{ECOOP}
\EventYear{2017}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{1}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\renewcommand{\rgn}{\pi}
\input{type-inference}
\input{misc}
\renewcommand{\rgn}{r}

@@

\clearpage

@@

\input{appendix}
\input{typesystem-proofs}
\input{fb-constraint-generation-extra}
\renewcommand{\rgn}{\myownthrowexception}
\input{typeinference-proofs}
\input{misc}

### 2017-01-13 paper/type-system.tex
Sec.3 edits complete.

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote objects contained in a single region, it
yeilds benefits that far outweigh the costs. In particular, it lets us
support region-polymorphic higher-order functions as class fields.
This allows, for example, a generic class, whose type parameters are
$a$ and $b$, to contain a region-polymorphic function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

The type rule for \C{new \RgnZ} expressions expects the \RgnZ class's
constructor to be called with a nullary function that returns a value
in its allocation context. 
% It enforces this by typechecking the body ($e$) of the function
% under an empty context containing nothing but the allocation context
% of the function. 
This ensures that the value returned by the function stores no
references to objects allocated elsewhere, including the top region
($\toprgn$), thus preventing cross-region references originating from
transferable regions. The body of the function might however create
new regions while execution, but this is not a problem as long as such
regions, and objects allocated in them, don't find their way into the
result of its evaluation.

@@

Consider a method $m$ with two region parameters - $\rho_0$ and
$\rho_1$ (i.e., $m\inang{\rho_0,\rho_1}(\dots)$). Suppose the method
immediately returns a function closure that contains a reference to
(an object in) $\rho_1$. Since function closures are always allocated
in the current allocation context, the closure is allocated in
$\rho_0$, the allocation context for the method body\footnote{
Fig.~\ref{fig:fb-staticsem-1} of the Appendix contains the method
well-formedness rule, which is on the similar lines as the lambda type
rule shown in Fig.~\ref{fig:fb-staticsem}}. Now, consider the
following use of the method:
\begin{codejava}
letregion R0 in
  let f = letregion R1 in m<R0,R1>()
  in f()
\end{codejava}
Observe that the first region parameter of \C{m} is instantiated with
\C{R0}, while the allocation context for the method call is \C{R1}.
The function \C{f} returned by \C{m} stores a reference to \C{R1},
which is unsafe when \C{f} is finally called. The type system
fortunately disallows this scenario by enforcing certain restrictions
during method and function applications as described above.
% The type rule for the \C{new} expression is a straightforward
% adaptation of corresponding FGJ rule to the the region type setting.

@@

@@

Fig.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} in the appendix
defines a small-step operational semantics for \fbname via a
four-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
Operational semantics replaces abstract region annotations ($\pi$ and
$\rho$) with \emph{locations} ($\loc$), where each \emph{location}
abstracts all the memory locations associated with a region (i.e.,
each memory region is associated with a single location). The
typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation relates an expression ($e$) and a typestate map ($\mem$) to a
reduced expression ($e'$) and an updated typestate map ($\mem'$). The
semantics gets ``stuck'' if $e$ attempts to access an object whose
allocation region is not present in $\rhoenv$, or if $e$ tries to
\C{open} a transferable region that is not mapped to an appropriate
typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
operation on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

Replacing $\pi$'s and $\rho$'s with $\loc$'s at runtime results in a
new class of expressions that are not captured in
Fig.~\ref{fig:fb-syntax}. Appendix contains full syntax that includes
such expressions. To help state the type safety theorem, we define the
syntactic class of (runtime) values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\;
    B\inang{\tbar}\inang{\overline{\loc}}(\bar{v}) \ALT
\lambdaexp{\loc}{\rhobar\,|\,\phi}{\taubar \; \xbar}{e} \ALT
\C{new}\; \RgnZT{\toploc\loc}(v)
%  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
The first two forms are obtained by using locations for region
annotations in $\C{new}$ and lambda expressions. The last form is the
value that $\C{new}\;\RgnZ$ expressions gets evaluated to; the
location $\toploc$ stands for the region $\toprgn$, and $\loc$ is the
location of the newly allocated transferable region.  The following
type safety theorem shows that a well-typed program will never attempt
to dereference an ``invalid'' reference (a reference to an object in a
region that has been transferred or freed):
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{\Delta}{(e,\mem)}{\invalidexn}$), or there exists an
$e'$ and a $\mem'$ such that $\consistent(\Delta,\mem')$ and
$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$ and
$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
\end{theorem}
The relation $\consistent$ relates $\Delta$ and $\mem$ only if both
make consistent assumptions about liveness of regions. Since $\Delta$
is consistent with both $\mem$ and $\mem'$, the theorem also captures
the key property of operational semantics that no live region is ever
freed or transferred.

% Furthermore, we prove the following theorem about \FB, which, in
% conjunction with the type safety theorem, implies the safety of region
% transfers across address spaces:
% \begin{theorem}
% \emph{(\textbf{Transfer Safety})}
% \label{thm:fb-transfer-safety}
% $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\mem$, $\mem'$, $\phicx$,
% $\phicx'$, such that $\loc \not\in dom(\mem')$, if $\hastyp{\emptyA, \cdot}{\C{new}\;
% \RgnZT{\toploc\loc}(v)}{\RgnZT{\toprgn}}$, then
% $\hastyp{\rhoenv',\cdot,\phicx')}
% {\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
% \end{theorem}
% The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
% is well-typed under one context, then it is also well-typed under
% every other context, whose $\mem$ maps $\rgn_i$ to closed
% ($\CLOSED$) state.  Thus, a recipient of a transferable region only
% needs to add a binding for the region to its $\mem$ in order to
% preserve its type safety. 
% Notably, this result could not have been
% established if it were possible for a transferable region to contain
% references to objects outside the region.

>>>>

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote the types of objects contained in a
single region, it yeilds benefits that far outweigh the costs. In
particular, it lets us support region-polymorphic higher-order
functions as class fields.  This allows, for example, a generic class,
whose type parameters are $a$ and $b$, to contain a region-polymorphic
function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

The type rule for \C{new \RgnZ} expressions expects the \RgnZ class's
constructor to be called with a nullary function that returns a value
in its allocation context.  This ensures that the value returned by
the function stores no references to objects allocated elsewhere,
including the top region ($\toprgn$), thus preventing cross-region
references originating from transferable regions. The body of the
function might however create new regions while execution, but this is
not a problem as long as such regions, and objects allocated in them,
don't find their way into the result of its evaluation.

% \begin{figure}[!t]
%   \input{fb-morewfrules}
%   \caption{\fbname: method and class well-formedness rules}
%   \label{fig:fb-morewfrules}
% \end{figure}

@@

Consider a method $m$ with two region parameters - $\rho_0$ and
$\rho_1$ (i.e., $m\inang{\rho_0,\rho_1}(\dots)$). Suppose the method
immediately returns a function closure that contains a reference to
(an object in) $\rho_1$. Since function closures are always allocated
in the current allocation context, the closure is allocated in
$\rho_0$, the allocation context for the method body (the rule for
methods discussed below). Now, consider the following use of the
method:
\begin{codejava}
letregion R0 in
  let f = letregion R1 in m<R0,R1>()
  in f()
\end{codejava}
Observe that the first region parameter of \C{m} is instantiated with
\C{R0}, while the allocation context for the method call is \C{R1}.
The function \C{f} returned by \C{m} stores a reference to \C{R1},
which is unsafe when \C{f} is finally called. The type system
fortunately disallows this scenario by enforcing certain restrictions
during method and function applications as described above.
% The type rule for the \C{new} expression is a straightforward
% adaptation of corresponding FGJ rule to the the region type setting.

@@

The method well-formedness rule makes use of an auxiliary definition
$\override$ that judges whether the current method is a valid
overriding of any eponymous method from the super class. The type
environment is extended with a binding that binds the $\thisZ$ keyword
to the type of the current class. Note that the the type of the
current method as accessed via $\thisZ$ (i.e., $\thisZ.m$) is
region-parametric, thus admitting region-polymorphic recursion (i.e.,
recursive call to $m$ can have different region arguments than $m$).

@@

\input{fb-syntax-extra}

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
additional langauge constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

Operational semantics of \FB defines a four-place small-step reduction
relation of the form shown below:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
The typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation relates an expression ($e$) and a typestate map ($\mem$) to a
reduced expression ($e'$) and an updated typestate map ($\mem'$). The
semantics gets ``stuck'' if $e$ attempts to access an object whose
allocation region is not present in $\rhoenv$, or if $e$ tries to
\C{open} a transferable region that is not mapped to an appropriate
typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
operation on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$). An illustrative subset of operational semantics
that formalize the intuitions desribed above is shown in
Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix
contains rest of the rules.

\input{fb-opsem}

To help state the type safety theorem, we define the syntactic class
of (runtime) values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\;
    B\inang{\tbar}\inang{\overline{\loc}}(\bar{v}) \ALT
\lambdaexp{\loc}{\rhobar\,|\,\phi}{\taubar \; \xbar}{e} \ALT
\C{new}\; \RgnZT{\toploc\loc}(v)
%  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
The first two forms are obtained by using locations for region
annotations in $\C{new}$ and lambda expressions. The last form is the
value that $\C{new}\;\RgnZ$ expressions gets evaluated to; the
location $\toploc$ stands for the region $\toprgn$, and $\loc$ is the
location of the newly allocated transferable region.  The following
type safety theorem shows that a well-typed program will never attempt
to dereference an ``invalid'' reference (a reference to an object in a
region that has been transferred or freed):
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{\Delta}{(e,\mem)}{\invalidexn}$), or there exists an
$e'$ and a $\mem'$ such that $\consistent(\Delta,\mem')$ and
$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$ and
$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
\end{theorem}
The relation $\consistent$ relates $\Delta$ and $\mem$ only if both
make consistent assumptions about liveness of regions. Since $\Delta$
is consistent with both $\mem$ and $\mem'$, the theorem also captures
the key property of operational semantics that no live region is ever
freed or transferred.

{\bf A note on region handlers}. \FB stores the region handlers in a
distinguished top region, which abstracts the GC heap. Type system
prevents transferable regions from storing references into the GC
heap. A static region however can store references to region handles,
which need to be taken into account while GC'ing the heap.  Traversing
the region memory to identify references into the GC heap
unfortunately defeats the purpose of regions. A solution is to disable
collecting region handles as long as any static region is open. Since
static regions are intended to store temporary objects while
populating a dynamic region, their lifetimes are short enough to let
region handles to be GC’d.
